---
title: "Blank Triplicate Test"
author: "Daniel Höhn"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

harmonicMean = function(vec){
  return(length(vec) / sum(vec^-1))
}
```

```{r triplicate 2, include=FALSE}
### feature lists ###
repli1 = read_csv("~/Work/master/QC_1_positive_features.csv")
repli1$type = "rep01"
repli2 = read_csv("~/Work/master/QC_2_positive_features.csv")
repli2$type = "rep02"
repli3 = read_csv("~/Work/master/QC_3_positive_features.csv")
repli3$type = "rep03"
repli4 = read_csv("~/Work/master/QC_4_positive_features.csv")
repli4$type = "rep04"
repli5 = read_csv("~/Work/master/QC_5_positive_features.csv")
repli5$type = "rep05"
repli6 = read_csv("~/Work/master/QC_6_positive_features.csv")
repli6$type = "rep06"
repli7 = read_csv("~/Work/master/QC_7_positive_features.csv")
repli7$type = "rep07"
repli8 = read_csv("~/Work/master/QC_8_positive_features.csv")
repli8$type = "rep08"
repli9 = read_csv("~/Work/master/QC_9_positive_features.csv")
repli9$type = "rep09"
repli10 = read_csv("~/Work/master/QC_10_positive_features.csv")
repli10$type = "rep10"
repli11 = read_csv("~/Work/master/QC_11_positive_features.csv")
repli11$type = "rep11"
repli12 = read_csv("~/Work/master/QC_12_positive_features.csv")
repli12$type = "rep12"
repli13 = read_csv("~/Work/master/QC_13_positive_features.csv")
repli13$type = "rep13"

### component regressions ###
comp1 = read_csv("~/Work/master/QC_1_positive_components.csv")
comp1$type = "rep01"
comp2 = read_csv("~/Work/master/QC_2_positive_components.csv")
comp2$type = "rep02"
comp3 = read_csv("~/Work/master/QC_3_positive_components.csv")
comp3$type = "rep03"
comp4 = read_csv("~/Work/master/QC_4_positive_components.csv")
comp4$type = "rep04"
comp5 = read_csv("~/Work/master/QC_5_positive_components.csv")
comp5$type = "rep05"
comp6 = read_csv("~/Work/master/QC_6_positive_components.csv")
comp6$type = "rep06"
comp7 = read_csv("~/Work/master/QC_7_positive_components.csv")
comp7$type = "rep07"
comp8 = read_csv("~/Work/master/QC_8_positive_components.csv")
comp8$type = "rep08"
comp9 = read_csv("~/Work/master/QC_9_positive_components.csv")
comp9$type = "rep09"
comp10 = read_csv("~/Work/master/QC_10_positive_components.csv")
comp10$type = "rep10"
comp11 = read_csv("~/Work/master/QC_11_positive_components.csv")
comp11$type = "rep11"
comp12 = read_csv("~/Work/master/QC_12_positive_components.csv")
comp12$type = "rep12"
comp13 = read_csv("~/Work/master/QC_13_positive_components.csv")
comp13$type = "rep13"
```



## Feature Grouping of ten replicate measurements
Data: Ten replicate measurements of standard compounds in water
Hypothesis: Features that can be replicated will differ in characteristic properties from those that can not be replicated
at all. 
Groups were formed by first sorting by mz and separating the groups every time that the difference between two adjacent
features exceeded the mz uncertainty. Within these groups, features were sorted by RT. The RT tolerance was set by taking a 
sample of groups of size ten with all replicates represented and taking approximately the largest differences between
RTS in those groups (four seconds). All groups were then split if that distance was exceeded. Between adjacent members
In total, 1675 of 89517 Features were in groups with more than ten members. These were removed from the dataset, along
with all other groups that contained more than one feature from each replicate.
```{r ten reps}
combo = rbind(repli1, repli2, repli3, repli4, repli5, repli6, repli7, repli8, repli9, repli10, repli11, repli12, repli13)

# combo = combo[which(combo$retentionTime > )]

  size = length(combo$ID)
  combo = combo[order(combo$mz), ]
  combo$mzdiff = c(combo$mz[-1], 0) - combo$mz
  combo$mzdiff[length(combo$mz)] = 0
  
  limitbreak = combo$mzdiff > combo$mzUncertainty # * 10e-6 * 2
  limitbreak = c(F, limitbreak[1:(length(limitbreak) - 1)])
  
  combo$group = cumsum(limitbreak) + 1
  combo = combo[order(combo$group, combo$retentionTime), ]
  
  limitbreak = limitbreak |
    (c(F, (
      c(combo$retentionTime[-1], 0) - combo$retentionTime
    ) > 4)[-length(limitbreak)])
  combo$group = cumsum(limitbreak) + 1
  
  combo = combo %>% group_by(group) %>% mutate(groupsize = n(), U_members = length(unique(type)))
  
  tmp = tabulate(combo$groupsize)
  tmp = tmp / c(1:max(combo$groupsize))
 barplot(tmp)
 
 combo$groupsize[which(combo$groupsize > 13)] = 13
 
 ggplot(combo)+
   geom_bar(aes(x = as.factor(groupsize), fill = type), position = "fill")+
   scale_x_discrete(name = "Replicates Found")+
   scale_y_continuous(name = "Percentage of Group Members")+
   labs(fill = "Replicate ID")
 
 
 # combo = combo[which(combo$groupsize == combo$U_members | combo$groupsize == 100), ]
 combo$points = combo$binIdxEnd - combo$binIdxStart + 1
 
 combo$isComp = combo$CompID > 0
```
Observations:
The majority of all features could be replicated, but a large portion was only found in one measurement.
There is no observable bias towards one group in any of the replicate groups, including those that contained
more than ten points.

```{r component replicability}
# how likely is it that a feature with an assigned component is found in all replicates?
ggplot(combo[which(combo$isComp),])+
   geom_bar(aes(x = as.factor(groupsize), fill = type))+
   scale_x_discrete(name = "Replicates Found")+
   scale_y_continuous(name = "Features Assigned to Components")+
   labs(fill = "Replicate ID")+
  theme(axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15))

ggplot(combo[which(combo$isComp),])+
  geom_boxplot(aes(x = log10(area), y = as.factor(groupsize)))+
   scale_y_discrete(name = "Replicates Found")+
   scale_x_continuous(name = "Feature Intensity (Log10)")+
  theme(axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15))

compCounts = combo %>% group_by(groupsize) %>% summarise(sum_y = sum(isComp), sum_n = sum(!isComp))

ggplot()+
  geom_point(aes(x = as.factor(1:max(compCounts$groupsize)), y = 100 * compCounts$sum_y / (compCounts$sum_n + compCounts$sum_y)), size = 4)+
   scale_x_discrete(name = "Replicates Found")+
   scale_y_continuous(name = "% of Component Features")+
   labs(fill = "Replicate ID")+
  theme(axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15))

ggplot(combo)+ # this is interesting, quantiles for groups of size four are much greater regarding area
  geom_boxplot(aes(x = as.factor(groupsize), y = area), outliers = F)+
   scale_x_discrete(name = "Replicates Found")+
   scale_y_continuous(name = "Area")+
  theme(axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15))

ggplot(combo[which(combo$isComp),])+ # this is interesting, quantiles for groups of size four are much greater regarding area
  geom_boxplot(aes(x = as.factor(groupsize), y = area), outliers = F)+
   scale_x_discrete(name = "Replicates Found")+
   scale_y_continuous(name = "Area")+
  theme(axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15))
```



```{r visualise}
options(dplyr.summarise.inform = FALSE)
# uncertainty is higher for group 4
ggplot(combo)+
  geom_boxplot(aes(x = as.factor(groupsize), y = mzUncertainty / mz), outliers = F)

ggplot(combo)+
  geom_boxplot(aes(x = as.factor(groupsize), y = mz))

ggplot(combo)+
  geom_boxplot(aes(x = as.factor(groupsize), y = retentionTime))

ggplot(combo)+
  geom_boxplot(aes(x = as.factor(groupsize), y = DQSC))

ggplot(combo)+
  geom_boxplot(aes(x = as.factor(groupsize), y = abs(DQSF)))

ggplot(combo)+ # this is interesting, quantiles for groups of size four are much greater regarding area
  geom_boxplot(aes(x = as.factor(groupsize), y = area), outliers = F)

ggplot(combo)+
   geom_bar(aes(x = as.factor(groupsize), fill = as.factor(competitors)), position = "fill")

ggplot(combo)+
   geom_bar(aes(x = as.factor(groupsize), fill = as.factor(scale)), position = "fill")

combo = combo[which(combo$groupsize == combo$U_members), ]
groupsizes = tabulate(combo$groupsize)
# counts_DF_c = as.data.frame(table(combo[,c(15,31)]))
# counts_DF_c = counts_DF_c[which(counts_DF_c$Freq != 0),]
# ggplot(counts_DF_c)+
#   geom_point(aes(y = (scale), x = (points), size = Freq))

# group_inter = as.data.frame(table(combo[,c(29, 16)]))
# group_inter$relative = group_inter$Freq / rep(groupsizes, length(unique(combo$interpolations))) # starts at 0
group_compete = as.data.frame(table(combo[,c(29, 17)]))
group_compete$relative = group_compete$Freq / rep(groupsizes, max(combo$competitors) + 1) # starts at 0
group_c = combo %>% group_by(groupsize, competitors) %>% 
  summarise(mean_width = mean(area / height), meanDQSF = mean(abs(DQSF)), meanMZ = mean(mz), 
            meanRT = mean(retentionTime), meanPoints = mean(binIdxEnd - binIdxStart + 1), meanArea = mean(area))
group_i = combo %>% group_by(groupsize, interpolations) %>% 
  summarise(mean_width = mean(area / height), meanDQSF = mean(abs(DQSF)), meanMZ = harmonicMean(mz), 
            meanRT = mean(retentionTime), meanPoints = mean(binIdxEnd - binIdxStart + 1), meanArea = harmonicMean(area))
group_s = combo %>% group_by(groupsize, scale) %>% 
  summarise(mean_width = mean(area / height), meanDQSF = mean(abs(DQSF)), meanMZ = harmonicMean(mz), 
            meanRT = mean(retentionTime), meanPoints = mean(binIdxEnd - binIdxStart + 1), meanArea = harmonicMean(area))

# not being replicable means higher likelihood of having interpolations
# ggplot(group_inter)+
#  geom_point(aes(x = interpolations, y = relative, colour = groupsize))

# features with no competitors are overrepresented in no_replicate
ggplot(group_compete)+
  geom_point(aes(x = competitors, y = relative, colour = as.factor(groupsize)))

ggplot(group_c)+
  geom_point(aes(x = as.factor(competitors), y = mean_width, colour = as.factor(groupsize)))

ggplot(group_s)+
  geom_point(aes(x = as.factor(scale), y = mean_width, colour = as.factor(groupsize)))

ggplot(group_c)+
  geom_point(aes(x = as.factor(competitors), y = meanDQSF, colour = as.factor(groupsize)))

ggplot(group_s)+
  geom_point(aes(x = as.factor(scale), y = meanDQSF, colour = as.factor(groupsize)))

ggplot(group_c)+
  geom_point(aes(x = as.factor(competitors), y = meanMZ, colour = as.factor(groupsize)))

ggplot(group_s)+
  geom_point(aes(x = as.factor(scale), y = meanMZ, colour = as.factor(groupsize)))

# ggplot(group_c)+ # effects only start showing for low point numbers
#   geom_point(aes(x = as.factor(competitors), y = meanRT, colour = as.factor(groupsize)))

ggplot(group_s)+
  geom_point(aes(x = as.factor(scale), y = meanRT, colour = as.factor(groupsize)))

ggplot(group_c)+ # no real information gain
  geom_point(aes(x = as.factor(competitors), y = meanPoints, colour = as.factor(groupsize)))

# ggplot(group_s)+
#   geom_point(aes(x = as.factor(scale), y = meanPoints, colour = as.factor(groupsize))) # just a straight line

ggplot(group_c)+ 
  geom_point(aes(x = as.factor(competitors), y = log(meanArea), colour = as.factor(groupsize)))

ggplot(group_s)+
  geom_point(aes(x = as.factor(scale), y = meanArea, colour = as.factor(groupsize)))

```

Discuss: Why are regressions with an even number of points favoured? 
Observation: When an even number of points is found in the regression, it is less likely for a point to be interpolated.
             When the regression covers an uneven number of real points, it is very likely that one point is interpolated.
Probably related to fixed uneven size of regression window, but that doesn't explain why even numbers are favoured



  